Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-07-25 paper/paper.tex
Cleaned up universal relations subsection of EOB section.

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}
Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.

@@

The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}. The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}. We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics. When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.

Let us now discuss the construction of the full waveform. In this work we only consider the dominant mode, $(\ell,m)=(2,2)$. The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$. Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum. Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum. Then we put
\begin{equation}
t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
\end{equation}
We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency. In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest. Explicitly,
\begin{equation}
|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
\end{equation}
where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger. This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}. The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
\begin{align}
\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
\end{align}
where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$. A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations. Finally, the full waveform reads
\begin{align}
h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
\end{align}

@@

As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless components of the spins $\boldsymbol{S}_{A,B}$ along the Newtonian angular momentum ${\boldsymbol{L}}_N$, $-1 \leq \chi_{A,B}=(\boldsymbol{S}_{A,B}\cdot \hat{\boldsymbol{L}}_N)/M_{A,B}^2 \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature. In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$. While this relation has been fitted to NR data only for $1 \leq \Lambda_2^{A,B}\leq10^6$, we use it also outside such range. However, this formula is divergent at small values of $\Lambda_2^{A,B}$, thus in such regime we replace it with a polynomial function that vanishes at 0, equals the universal relation at $\Lambda_2^{A,B}=10^{-2}$, and best-fits it for $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$ Explicitly, putting $\zeta_{A,B}=\frac{\log\Lambda^{A,B}_2}{10}$,
\begin{widetext}
\begin{equation}
\Lambda_3^{A,B}=\left\{\begin{array}{ll}
\Lambda_2^{A,B}\left[0.441 - 34.6\,\Lambda_2^{A,B} + 1760(\Lambda_2^{A,B})^2\right]\,,&\textrm{if}\quad0\leq\Lambda_2^{A,B}\leq10^{-2}\\
&\\
\exp{\left[-1.15 +11.8\,\zeta_{A,B}+2.51\,\zeta_{A,B}^2-1.31\,\zeta_{A,B}^3+0.252\,\zeta_{A,B}^4\right]}\,,&\textrm{otherwise}\,.
\end{array}\right.
\end{equation}
\end{widetext}
Eq. (3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent. Putting $\xi_{A,B}=\log\Lambda_2^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $0\leq \xi_{A,B} \leq 9$), we set
\begin{align}
M_{A,B}\omega_{02}^{A,B}&=
0.182 -6.836\times10^{-3}\xi_{A,B}\nonumber\\
&-4.196\times10^{-3}\xi_{A,B}^2+5.215\times10^{-4}\xi_{A,B}^3\nonumber\\
&-1.857\times10^{-5}\xi_{A,B}^4\,;\label{URomega2}
\end{align}
outside this range, we set $M_{A,B}\omega_{02}^{A,B}=0.182$ if $\xi_{A,B} < 1$ and $M_{A,B}\omega_{02}^{A,B}=0.161$ if $\xi_{A,B} >9$, that is Eq.~(\ref{URomega2}) evaluated for $\xi_{A,B}=1$ and $\xi_{A,B}=9$, respectively. Moreover, putting $\Upsilon_{A,B}=\log\Lambda_3^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $-1 \leq \Upsilon_{A,B} \leq 10$), we set
\begin{align}
M_{A,B}\omega_{03}^{A,B}&=
\,0.2245 -1.5\times10^{-2}\Upsilon_{A,B}\nonumber\\
&-1.412\times10^{-3}\Upsilon_{A,B}^2+1.832\times10^{-4}\Upsilon_{A,B}^3\nonumber\\
&-5.561\times10^{-6}\Upsilon_{A,B}^4\,;\label{URomega3}
\end{align}
outside this range, we set $M_{A,B}\omega_{03}^{A,B}=0.2379$ if $\Upsilon_{A,B} < -1$ and $M_{A,B}\omega_{03}^{A,B}=0.1165$ if $\Upsilon_{A,B} >10$, that is Eq.~(\ref{URomega3}) evaluated for $\Upsilon_{A,B}=-1$ and $\Upsilon_{A,B}=10$, respectively.
Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$. We note that this list does not include the total mass $M$ of the system. Of course, the point-mass part of the model is scale-invariant. As to the inspiral-plunge phase, in this model tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the post-Newtonian analysis of tidal effects~\cite{Flanagan:2007ix}. Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either, although a more sophisticated description will have to include such dependence to account for the phenomenology that is observed in NR simulations of this phase.

\red{A notable effect that has not been included is the spin induced quadrupole moment of each NS [[Poisson etc.]]. The impact on detection and parameter estimation is discussed in [[HarryHinderer]].}

>>>>

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}

\subsection{Inspiral-plunge waveform}

Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion---be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.

Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.

@@

Let us now discuss the construction of the full waveform. The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.

The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.

The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$. Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum. Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum. Then we put
\begin{equation}
t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
\end{equation}
In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.

%We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics. When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.

\subsection{Post-merger waveform}

\red{[[Mention that this is just a tapering and does not include post-merger oscillations seen in numerical simulations.]]}

We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency. In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest. Explicitly,
\begin{equation}
|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
\end{equation}
where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger. This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}. The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
\begin{align}
\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
\end{align}
where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$. A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations. Finally, the full waveform reads
\begin{align}
h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
\end{align}

@@

\subsection{Approximation of higher-order tidal effects}

As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we exploit nearly EOS-independent fits for the higher-order effects in terms of $\Lambda_2$ usually referred to as universal relations. In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} provides a fit for $\Lambda_3$ as a function of $\Lambda_2$ with an accuracy of about 10\%, depending on the EOS. This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^6$. This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that vanishes at $\Lambda_2 = 0$. We require this extension to be continuous and smooth at $\Lambda_2 = 10^{-2}$, and to fit the universal relation in the range $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$. The exact relation we use is
\begin{equation}
\Lambda_3 = \left\{\begin{array}{ll}
\Lambda_2 (f + g \Lambda_2 + h \Lambda_2^2)\, , & 0 \leq \Lambda_2 \leq 10^{-2} \\
e^{a + b \xi + c \xi^2 + d \xi^3 + e \xi^4}\, , & \Lambda_2>10^{-2}
\end{array}\right.,
\end{equation}
where $\xi = \ln\Lambda_2$, $\{a, b, c, d, e\} = \{-1.15, 1.18, 2.51\times 10^{-2}, -1.31\times 10^{-3}, 2.52\times 10^{-5}\}$ from Ref.~\cite{Yagi:2013sva}, and $\{f, g, h\}=\{0.441, -34.6, 1760\}$.

Eq. (3.5) of Ref.~\cite{Chan:2014kua} gives relations for $\omega_{02}$ as a function of $\Lambda_2$ and $\omega_{03}$ as a function of $\Lambda_3$ to within a few percent error. For $\omega_{02}$, the fitting range used by Ref.~\cite{Chan:2014kua} was $1\leq \xi \leq 9$, and outside this range we require continuity. The relation we use is then
\begin{equation}
\omega_{02} = \left\{\begin{array}{ll}
f\, , & \xi < 1 \\
a + b\xi + c\xi^2 + d\xi^3 + e\xi^4\, , & 1 \le \xi \le 9 \\
g\, , & \xi > 9
\end{array}\right.,
\end{equation}
where $\{a, b, c, d, e\} = \{0.182, -6.836\times10^{-3}, -4.196\times10^{-3}, 5.215\times10^{-4}, -1.857\times10^{-5}\}$ from Ref.~\cite{Chan:2014kua}, and $\{f, g\}=\{0.182, 0.161\}$. For $\omega_{03}$, the fit is given in terms of $\Upsilon=\ln\Lambda_3$ in the range $-1\leq \Upsilon \leq 10$, and outside this range we require continuity. The relation we use is then
\begin{equation}
\omega_{03} = \left\{\begin{array}{ll}
f\, , & \Upsilon < -1 \\
a + b\Upsilon + c\Upsilon^2 + d\Upsilon^3 + e\Upsilon^4\, , & -1 \le \Upsilon \le 10 \\
g\, , & \Upsilon > 10
\end{array}\right.,
\end{equation}
where $\{a, b, c, d, e\} = \{0.2245, -1.5\times10^{-2}, -1.412\times10^{-3}, 1.832\times10^{-4}, -5.561\times10^{-6}\}$ and $\{f, g\}=\{0.2379, 0.1165\}$.

With these relations the waveform only depends on 5 intrinsic parameters $\mathbf{x}=\{q, S_A, S_B, \Lambda_2^A,\Lambda_2^B\}$. We note that this list does not include the total mass $M$ of the system. The point-mass part of the model is scale-invariant, and the tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$ and $\Lambda_2^{A,B}$. Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either. It is not clear at what point in the transition from the inspiral to the post-merger this approximation breaks down. For example, the frequency of post-merger oscillations depends sensitively on the total mass of the hypermassive remnant. A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.

%%% ________________________________________________________
%\section{Aligned-spin, dynamical tides EOB model}
%Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. 
%
%In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
%\begin{equation}
%\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
%\end{equation}
%where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
%\begin{widetext}
%\begin{align}
% \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
% &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
%\end{align}
%\end{widetext}
%where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies. In particular, the dimensionless dynamical tidal polarizability reads
%\begin{equation}
%\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
%\end{equation}
%where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
%\begin{equation}
%\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
%\end{equation}
%with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$. 
%
%Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
%\begin{equation}
%h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
%\end{equation} 
%where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
%
%The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}. The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}. We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics. When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case. 
%
%Let us now discuss the construction of the full waveform. In this work we only consider the dominant mode, $(\ell,m)=(2,2)$. The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$. Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum. Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum. Then we put
%\begin{equation}
%t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
%\end{equation}
%We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency. In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest. Explicitly,
%\begin{equation}
%|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
%\end{equation}
%where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger. This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}. The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
%\begin{align}
%\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
%&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
%\end{align}
%where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$. A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations. Finally, the full waveform reads
%\begin{align}
%h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
%&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
%\end{align}
%
%Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state. Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit. Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
%
%We note that, to date, this is the only tidal EOB model that includes the effect of spin. More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs. The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning. This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
%
%As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless components of the spins $\boldsymbol{S}_{A,B}$ along the Newtonian angular momentum ${\boldsymbol{L}}_N$, $-1 \leq \chi_{A,B}=(\boldsymbol{S}_{A,B}\cdot \hat{\boldsymbol{L}}_N)/M_{A,B}^2 \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature. In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$. While this relation has been fitted to NR data only for $1 \leq \Lambda_2^{A,B}\leq10^6$, we use it also outside such range. However, this formula is divergent at small values of $\Lambda_2^{A,B}$, thus in such regime we replace it with a polynomial function that vanishes at 0, equals the universal relation at $\Lambda_2^{A,B}=10^{-2}$, and best-fits it for $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$ Explicitly, putting $\zeta_{A,B}=\frac{\log\Lambda^{A,B}_2}{10}$,
%\begin{widetext}
%\begin{equation}
%\Lambda_3^{A,B}=\left\{\begin{array}{ll}
